---
title: "Compare NCBI and CAMDA species"
author: "Mariana Amaro"
format: html
editor: visual
---

This script was used to compare CAMDA database and NCBI data of species

1.  Charge R libraries

```{r}
# This are the libraries that we need to charge
library(dplyr)
library(ggplot2)
library(stringr)
library(tidyr)
library(viridis)
library(readxl)


```

2.  Upload the database

```{r}
#file routes of the database
#training_dataset=CAMDA sra_metadata=NCBI
training_dataset <- read_excel("C:/Users/52462/OneDrive/Documentos/CAMDA2025/CAMDA25/training_dataset.xlsx")
sra_metadata <- read_excel("C:/Users/52462/OneDrive/Documentos/CAMDA2025/CAMDA25/sra-metadata.xlsx")
```

3.  Change the titles if it's necessary in the species column

```{r}

# 1. Procesar columnas de especie (juntar genus y species)
training_dataset <- training_dataset %>%
  mutate(
    accession = as.character(accession),
    full_name = str_squish(paste(genus, species)),
    training_species = word(full_name, 1, 2)
  ) %>%
  select(accession, training_species)
```

```{r}

# 2. Procesar ScientificName (quedarse con 2 primeras palabras)
sra_metadata <- sra_metadata %>%
  mutate(
    Run = as.character(Run),
    sra_species = word(ScientificName, 1, 2)
  ) %>%
  select(Run, sra_species)
```

4.  Union of both dataset per accession to identify the differences between them

```{r}
# 3. Unir datasets por accession/Run
comparison <- full_join(training_dataset, sra_metadata, 
                        by = c("accession" = "Run"))

# 4. Identificar diferencias
comparison <- comparison %>%
  mutate(
    match_status = case_when(
      is.na(training_species) | is.na(sra_species) ~ "Missing",
      training_species == sra_species ~ "Match",
      TRUE ~ "Mismatch"
    )
  )
```

5.   Check the accessions that are different in NCBI. This data represent the column "new_species" in the file ManualSpeciesUpdates .

```{R}
# 5. Ver accesiones con especies distintas
discrepancies <- comparison %>%
  filter(match_status == "Mismatch")

# Mostrar los primeros registros discrepantes
print(head(discrepancies, 10))

# Guardar tabla de discrepancias si lo deseas
write.csv(discrepancies, "discrepancias_especies.csv", row.names = FALSE)

```

6.  Unidentified species in NCBI

```{r}
# Resumen de NAs por columna
na_summary <- sra_metadata %>%
  summarise(across(everything(), ~ sum(is.na(.)))) %>%
  pivot_longer(everything(), names_to = "columna", values_to = "num_NA") %>%
  filter(num_NA > 0) %>%
  arrange(desc(num_NA))

print(na_summary)

# Número total de valores NA en todo el dataset
total_NA <- sum(is.na(sra_metadata))
cat("Número total de valores NA en sra_metadata:", total_NA, "\n")

# Mostrar todas las filas con al menos un NA (sin límite)
missing_rows <- sra_metadata %>%
  filter(if_any(everything(), is.na))

print(missing_rows)


```
